Street Network Representation and Nonplanarity in Urban Form Studies

### 2018-02-01 manuscript.tex
revise intro conclusion

Models of street networks underlie research in urban travel behavior, accessibility, design patterns, and morphology. These models are commonly defined as \enquote{planar,} meaning that they can be represented in two dimensions without any underpasses or overpasses. However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels. How well do planar models represent real-world street networks? What does a city's extent of nonplanarity reveal about its infrastructure and urban development? This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city. It develops new measures of a network's nonplanarity as indicators of infrastructure and urbanization. While some street networks are approximately planar, many are poorly-modeled by planar simplifications that inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.

@@

In the urban studies literature, street networks are often referred to as \enquote{planar} or \enquote{approximately planar,} meaning that they can be well-represented by a two-dimensional model that inherently precludes overpasses or underpasses. However, this claim has not been well studied or defined empirically. In the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels. This leads to two intertwined questions. First, how well do these planar graphs model urban street networks? Second, what does the extent to which a network is nonplanar tell us about a city's infrastructure and development?

This study tests the planarity of street networks around the world. It also presents two new measures of the extent of nonplanarity that can be generalized to other types of spatial networks. These indicators help describe the nature of the urban form and transportation infrastructure. Despite common claims in the literature that urban street networks are planar graphs, this study finds that they generally are nonplanar and that planar graphs poorly model the street networks of many urban centers. Further, the magnitude of this bias varies substantially across cities and urbanization types.

This paper first introduces the basics of graph theory relevant to urban studies, focusing on discussions in the research literature about street network planarity. Next it discusses the methods used to acquire and analyze the street networks in this study. Then it present the results of this analysis before concluding with a discussion of ramifications for street network research and urban form studies.

@@

If street networks can be sufficiently well-modeled by planar graphs, there are certain methodological benefits to doing so. Planar graphs offer computational simplicity and tractability. They enable simple polygonal analysis of city blocks and form \citep{fohl_non-planar_1996,barthelemy_paths_2017} as well as the $\alpha$ and $\gamma$ indices popular in road network analysis \citep{eppstein_studying_2008}. In mathematics, a bijection exists planar graphs and trees, and classifying planar graphs presents a trivial problem \citep{louf_typology_2014}. Planar graphs are easier to visualize and can be faster to run algorithms on \citep{liebers_planarizing_2001}. Accordingly, \citet[p.~3]{barthelemy_spatial_2011} argues that \enquote{planar spatial networks are the most important and most studies have focused on these examples}. But in contrast, \citet{masucci_random_2009} and \citet{masucci_limited_2013} argue that planar graphs remain a compelling research domain for urban scholars because they were understudied until recently for two reasons: they appear topologically trivial and planarity does not lend itself to certain popular graph-theoretic analyses. Discussing the open research area around street networks as planar graphs, \citet[p.~1]{viana_simplicity_2013} state, \enquote{there is still a lack of global, high-level metrics allowing to characterize their structure and geometrical patterns.}

@@

This study develops two new measures of the extent to which a spatial network is planar. Then it analyzes various city centers worldwide to better understand how well planar graphs model urban street networks and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization. Finally, it considers what these indicators suggest about the urban form and transportation infrastructure in different cities.

@@

Table \ref{tab:world_cities} lists the planarity indicators for these 50 city centers. Among the drivable street networks, only 20\% are formally planar. On average, they are 88.5\% spatially planar by the $\phi$ measure and 90\% by the $\lambda$ measure. The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 57\% in Moscow. The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles. On average across these networks, planar representations overcount intersections by 16\% and underestimate street segment lengths by 10\% (Equations \ref{eq:spr_overstates} and \ref{eq:elr_understates}).

Among walkable street networks, only 10\% are formally planar. On average, they are 92\% spatially planar by the $\phi$ measure and 91.5\% by the $\lambda$ measure. The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 67\% in Shanghai. The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 66\% in Shanghai. On average across these networks, planar representations overcount intersections by 9\% and underestimate street segment lengths by 8.5\%. Fewer walkable than drivable networks are formally planar, but on average these walking networks are slightly more spatially planar than the driving networks.

@@

However, these indicators' values vary across the city. To explore this statistical variation, Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network. Our samples' mean $\phi$ and $\lambda$ scores are reasonably close to the city-wide values. However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 63.7\% ($\lambda$) up to highs of 100\%. 67\% of the samples pass the formal planarity test; however, 63\% of the samples are at least somewhat spatially nonplanar (i.e., with $\phi < 1$).

@@

Street networks are frequently spatially nonplanar because they are embedded in three dimensions, not two: they have a (less-extensive) z-axis along with their x- and y-axes. But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88.5\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, they could often be described as \emph{approximately} planar. However, claiming that urban street networks universally are \enquote{planar,} or modeling them as such, misrepresents them in several ways:

@@

A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes. Contrary to some of the statements in the urban studies literature, our results suggest that it cannot be universally claimed that urban street networks are planar graphs --- formally or spatially. However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar and its implications for modeling.

@@

But planar graphs are often used in the literature to characterize urban form and morphology. Aside from routing, do planar graphs offer useful models for this type of research? Again, only in limited circumstances, such as the drivable networks in the three Italian cities we analyzed. Elsewhere, the results in Table \ref{tab:world_cities} showed how common urban form measures such as intersection counts are overstated by planar models (16\% on average in the drivable networks), while average street segment lengths are consequently understated (10\% on average in the drivable networks). Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} and Table \ref{tab:samples_city} demonstrated how the magnitude of bias varies across cities and modes of urbanization.

We might shoehorn real-world data into a chosen model and be able to perform a desired computation, but we lose the underlying ability to reason when the real world no longer obeys the rules and constraints of the chosen model. Although planar graphs offer computational tractability and allow for the polygonal analysis of urban blocks, they often model these street networks poorly when compared to nonplanar graphs.

@@

We might repurpose this discrepancy between planar and nonplanar representations of a street network to reveal something useful about urban form. The $\phi$ and $\lambda$ scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers. These US city centers feature more automobile-oriented roadways and are often bounded by grade-separated freeways. By contrast, these European city centers feature less three-dimensionality as their streets and paths tend to be at-grade. Examining how $\phi$ and $\lambda$ change over time in different cities could reveal the type and pace of urban development. Future research can further explore how these indicators correlate with other measures of urbanization, development, and era.

@@

Street networks are often defined and modeled as planar graphs in the urban studies literature despite typically being nonplanar in reality. This study tested the formal planarity of central urban street networks and developed two new measures of their spatial planarity: the Spatial Planarity Ratio $\phi$ and the Edge Length Ratio $\lambda$. Although planar models may be useful in limited circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths. Nonplanar graphs usually provide more faithful models, and $\phi$ and $\lambda$ can characterize urbanization in different cities, particularly through the transportation infrastructure's three-dimensionality.

>>>>

Models of street networks underlie research in urban travel behavior, accessibility, design patterns, and morphology. These models are commonly defined as \enquote{planar,} meaning that they can be represented in two dimensions without any underpasses or overpasses. However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels. Planar simplifications affect the results of real-world street network analysis. A city's extent of nonplanarity can describe its infrastructure and urban development. This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city. It develops new nonplanarity indicators of infrastructure and urbanization. While some street networks are approximately planar, we empirically quantify how planar simplifications can inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.

@@

The urban studies literature often defines street networks as \enquote{planar} or \enquote{approximately planar,} meaning they can be well-represented by a two-dimensional model that inherently precludes overpasses or underpasses. However, this claim has not been well studied or defined empirically. In the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels. How well do planar graphs model urban street networks? To what extent do convenient planar simplifications alter network analysis results? What does the extent to which a network is nonplanar tell us about a city's infrastructure and development?

This study tests the planarity of street networks around the world and demonstrates how planar simplifications alter the results of street network analysis. It also presents two new measures of the \enquote{extent of nonplanarity} that describe the urban form and transportation infrastructure and can be generalized to other types of spatial networks. Despite common assumptions in the literature that urban street networks are planar graphs, this study finds that they generally are nonplanar and empirically quantifies how they misrepresent the street networks of many urban centers. Further, the magnitude of this bias varies substantially across cities and urbanization types.

This paper first introduces the basics of graph theory relevant to urban studies, focusing on discussions in the research literature about street network planarity. Next it discusses the methods used to acquire and analyze the street networks in this study. Then it present the results of this analysis before concluding with a discussion of implications for street network research and urban form studies.

@@

If street networks can be sufficiently well-modeled by planar graphs, there are certain methodological benefits to doing so. Planar graphs offer computational simplicity and tractability. They enable simple polygonal analysis of city blocks and form \citep{fohl_non-planar_1996,barthelemy_paths_2017} as well as the \emph{alpha} and \emph{gamma} indices popular in road network analysis \citep{eppstein_studying_2008}. In mathematics, a bijection exists planar graphs and trees, and classifying planar graphs presents a trivial problem \citep{louf_typology_2014}. Planar graphs are easier to visualize and can be faster to run algorithms on \citep{liebers_planarizing_2001}. Accordingly, \citet[p.~3]{barthelemy_spatial_2011} argues that \enquote{planar spatial networks are the most important and most studies have focused on these examples}. But in contrast, \citet{masucci_random_2009} and \citet{masucci_limited_2013} argue that planar graphs remain a compelling research domain for urban scholars because they were understudied until recently for two reasons: they appear topologically trivial and planarity does not lend itself to certain popular graph-theoretic analyses. Discussing the open research area around street networks as planar graphs, \citet[p.~1]{viana_simplicity_2013} state, \enquote{there is still a lack of global, high-level metrics allowing to characterize their structure and geometrical patterns.}

@@

This study develops two new measures of the extent to which a spatial network is planar. It empirically analyzes various city centers worldwide to quantify how planar models impact street network analysis results and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization. Finally, it considers what these indicators suggest about the urban form and transportation infrastructure in different cities.

@@

Table \ref{tab:world_cities} demonstrates how planar simplifications impact street network analysis results in these 50 city centers. Among the drivable street networks, only 20\% are formally planar. On average, they are 88.5\% spatially planar by the $\phi$ measure and 90\% by the $\lambda$ measure. The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 57\% in Moscow. The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles. On average across these networks, planar models overcount intersections by 16\% and underestimate street segment lengths by 10\% (Equations \ref{eq:spr_overstates} and \ref{eq:elr_understates}).

Among walkable street networks, only 10\% are formally planar. On average, they are 92\% spatially planar by the $\phi$ measure and 91.5\% by the $\lambda$ measure. The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 67\% in Shanghai. The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 66\% in Shanghai. On average across these networks, planar models overcount intersections by 9\% and underestimate street segment lengths by 8.5\%. Fewer walkable than drivable networks are formally planar, but on average these walking networks are slightly more spatially planar than the driving networks.

@@

However, these indicators' values vary across the city. To explore this statistical variation, Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network. Our samples' mean $\phi$ and $\lambda$ scores are reasonably close to the city-wide values. However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 63.7\% ($\lambda$) up to highs of 100\%. 67\% of the samples pass the formal planarity test; however, 63\% of the samples are at least somewhat spatially nonplanar (i.e., $\phi < 1$).

@@

Street networks are frequently spatially nonplanar because they are embedded in three dimensions, not two. But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88.5\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, they could often be described as \emph{approximately} planar. However, claiming that urban street networks universally are \enquote{planar,} or modeling them as such, misrepresents them in several ways:

@@

A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes. These results suggest that it cannot be universally claimed that urban street networks are planar graphs --- formally or spatially. However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar. How do planar simplifications impact modeling and analysis?

@@

But aside from routing, planar graphs are often used to characterize urban form and morphology. Do they provide useful models for this type of research? Again, only in limited circumstances, such as the drivable networks in the three Italian cities we analyzed. Elsewhere, Table \ref{tab:world_cities} showed how the results of common urban form analyses such as intersection counts are overstated by planar models (16\% on average in the drivable networks), while average street segment lengths are consequently understated (10\% on average in the drivable networks). Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} and Table \ref{tab:samples_city} demonstrated how the magnitude of bias varies across cities and modes of urbanization.

We may be able to shoehorn real-world data into a chosen model and perform a desired computation, but we lose the underlying ability to reason when the real world no longer obeys the rules and constraints of the chosen model. Although planar simplifications offer algorithmic tractability and allow for the polygonal analysis of urban blocks, these results demonstrate how they can model urban street networks poorly and bias analytical results.

@@

We might repurpose this discrepancy between planar and nonplanar models of a street network to reveal something useful about urban form. The $\phi$ and $\lambda$ scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers. These US city centers feature more automobile-oriented roadways and are often bounded by grade-separated freeways. By contrast, these European city centers feature less three-dimensionality as their streets and paths tend to be at-grade. Examining how $\phi$ and $\lambda$ change over time in different cities could reveal the type and pace of urban development. Future research can further explore how these indicators correlate with other measures of urbanization, development, and era.

@@

Street networks are often defined and modeled as planar graphs in the urban studies literature despite typically being nonplanar in reality. This study tested the formal planarity of central urban street networks and  empirically demonstrated how planar simplifications affect analysis results. It developed two new measures of the extent to which they are nonplanar --- the Spatial Planarity Ratio $\phi$ and the Edge Length Ratio $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality. Although planar models may be useful in limited circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
